\documentclass[11pt]{article}

\usepackage[final]{acl}

\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{booktabs}

\title{Reproducing and Extending KELE: A Multi-Agent Framework for Structured Socratic Teaching with LLMs}

\author{Ulises Chavarria \\
  Santa Clara University \\
  \texttt{uchavarria@scu.edu} \\\And
  Maximilian Khan \\
  Santa Clara University \\
  \texttt{mjkhan@scu.edu} \\}

\begin{document}
\maketitle

\begin{abstract}
We reproduce and extend KELE \citep{peng2025kele}, a multi-agent framework for structured Socratic teaching with large language models. KELE decomposes Socratic dialogue into five teaching stages governed by SocRule, using a consultant--teacher architecture where one LLM selects teaching strategies and another generates responses. We reimplement the full pipeline, serve SocratTeachLLM locally via vLLM, and evaluate on the 681-dialogue test split of SocratDataset. Our primary reproduction with GPT-4o as live consultant achieves ROUGE-2 of 26.04 and BLEU-4 of 19.60; additional fully-local runs with open-weight consultants on diverse hardware (NVIDIA V100, AMD R9700) reveal that ROUGE/BLEU scores are stable across consultant quality while state accuracy degrades sharply, identifying structured state classification as the key bottleneck. We propose three improvements: a hierarchical supervised classifier to replace the LLM consultant, a stronger teacher base model, and retrieval-augmented generation to close the surface-form gap. Code and data are available at \url{https://github.com/ulises-c/csen-346} and \url{https://huggingface.co/ulises-c/SocratTeachLLM}.
\end{abstract}

%% ============================================================
%% SECTION 1 — INTRODUCTION
%% ============================================================
\section{Introduction}

Dialogue plays a central role in shaping learning outcomes. Traditional knowledge-imparting instruction places students in a passive role, delivering answers directly without engaging deeper reasoning \citep{johnston1994teaching}. Socratic teaching offers an alternative: through structured questioning, learners are guided to actively construct knowledge and discover answers themselves \citep{seeskin1987socratic, chang1998socratic}. Research consistently shows that Socratic dialogue promotes cognitive development and deeper understanding \citep{knezic2010socratic}, yet its reliance on highly skilled instructors makes it difficult to scale in practice.

Recent advances in large language models (LLMs) have opened a path toward automating Socratic instruction. LLMs are increasingly applied in educational settings---student Q\&A, lesson planning, and adaptive tutoring \citep{kasneci2023chatgpt, gan2023llmedu}---and several systems have explored Socratic-style dialogue generation, including prompt-based approaches \citep{zhang2024spl, alhossami2024socratic} and fine-tuned models such as SocraticLM \citep{liu2024socraticLM} and EduChat \citep{dan2023educhat}. However, these efforts generally focus on surface-level heuristic questioning while neglecting the structured, multi-stage progression that characterizes effective Socratic pedagogy.

\citet{peng2025kele} address this gap with KELE (Knowledge-Enlightened Learning Enhanced by LLMs), a multi-agent framework that decomposes Socratic teaching into five hierarchical stages---student questioning, concept probing, inductive reasoning, rule construction, and teacher summary---governed by a formal rule system called SocRule containing 34 teaching strategies. The framework employs a ``consultant--teacher'' mechanism: a \emph{consultant} agent evaluates student cognitive state and selects the appropriate teaching strategy, while a \emph{teacher} agent generates the actual dialogue response. To power the teacher agent, the authors fine-tune GLM4-9B \citep{glm2024chatglm} using LoRA \citep{hu2022lora} on SocratDataset, a corpus of 6,803 dialogues with over 42,000 annotated turns, producing SocratTeachLLM---a model that outperforms GPT-4o across all nine Socratic teaching evaluation dimensions despite being significantly smaller.

In this work, we reproduce and extend the KELE framework with three goals:

\begin{enumerate}
    \item \textbf{Reproduction.} We reimplement the consultant--teacher pipeline, serve SocratTeachLLM locally via vLLM \citep{kwon2023vllm}, and use GPT-4o \citep{openai2024gpt4o} as the consultant to evaluate on the same 681-dialogue test split. We report ROUGE \citep{lin2004rouge}, BLEU \citep{papineni2002bleu}, and per-stage state accuracy to verify the original claims.
    \item \textbf{Diagnosis.} We analyze where the reproduction matches or diverges from the published results, with particular attention to the low mid-stage state accuracy (stages c and d below 6\%) that emerges from our evaluation.
    \item \textbf{Improvement.} We propose replacing the LLM-based consultant with a lightweight supervised classifier trained on SocratDataset's 42,000 state-labeled turns, targeting the system's weakest link: the consultant's ability to correctly identify student cognitive states during the complex middle stages of Socratic dialogue.
\end{enumerate}

Table~\ref{tab:reproduction} summarizes our baseline reproduction against the original results. Our system uses the same configuration as \citet{peng2025kele}: GPT-4o as the consultant agent and SocratTeachLLM as the teacher agent, evaluated on the 681-dialogue test split (10\% of SocratDataset). We include the paper's GPT-4o teacher baseline for additional context.

\begin{table}[t]
  \centering
  \small
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{@{}lcccc@{}}
    \toprule
    \textbf{System} & \textbf{R-1} & \textbf{R-2} & \textbf{R-L} & \textbf{BLEU-4} \\
    \midrule
    \multicolumn{5}{@{}l}{\emph{Peng et al.\ (2025) --- original results}} \\
    \;\; GPT-4o (teacher)         & 48.25 & 22.55 & 38.27 & 29.93 \\
    \;\; SocratTeachLLM           & \textbf{57.40} & \textbf{33.63} & \textbf{50.77} & \textbf{41.96} \\
    \midrule
    \multicolumn{5}{@{}l}{\emph{Our reproduction}} \\
    \;\; SocratTeachLLM           & 44.61 & 26.04 & 38.02 & 19.60 \\
    \midrule
    \multicolumn{5}{@{}l}{\emph{Delta (ours vs.\ original SocratTeachLLM)}} \\
    \;\;                          & --12.79 & --7.59 & --12.75 & --22.36 \\
    \bottomrule
  \end{tabular}%
  }
  \caption{Reproduction comparison on the 681-dialogue test split. R-1/2/L = ROUGE-1/2/L (F1). Both the original and our system use GPT-4o as the consultant and SocratTeachLLM as the teacher. Our ROUGE-L closely matches the paper's GPT-4o teacher baseline (38.02 vs.\ 38.27), and our ROUGE-2 exceeds it (26.04 vs.\ 22.55), confirming that SocratTeachLLM provides meaningful gains over a general-purpose teacher. The BLEU-4 gap likely stems from undocumented generation parameters in the original work.}
  \label{tab:reproduction}
\end{table}

%% ============================================================
%% SECTION 2 — RELATED WORK
%% ============================================================
\section{Related Work}

\subsection{Large Language Models in Education}

LLMs are rapidly being integrated across educational applications, from student Q\&A and personalized tutoring to lesson planning and intelligent feedback \citep{kasneci2023chatgpt, gan2023llmedu}. InstructGPT-style alignment through reinforcement learning from human feedback has enabled conversational agents that can adapt explanations to student needs \citep{chung2024scaling}. RetLLM-E \citep{mitra2024retllm} retrieves relevant course material to ground LLM responses for student forum questions, while LessonPlanner \citep{fan2024lessonplanner} assists novice teachers in generating pedagogically sound lesson plans. The CLASS framework \citep{sonkar2023class} builds intelligent tutoring systems grounded in learning science principles, guiding students step-by-step through interactive dialogue. These systems demonstrate the growing capacity of LLMs to serve as educational tools, though most operate in a single-agent paradigm without explicit pedagogical structure.

\subsection{Technology-Enhanced Socratic Teaching}

The Socratic method---teaching through guided questioning rather than direct instruction---has a long history in educational research \citep{seeskin1987socratic, chang1998socratic}. Technology-enhanced implementations range from early rule-based dialogue systems \citep{graesser2004autotutor} to modern LLM-powered approaches. \citet{alhossami2024socratic} demonstrate that LLMs can employ Socratic questioning to help students debug code, while \citet{shridhar2022mathword} show automatic generation of Socratic sub-questions for math word problems. The SPL system \citep{zhang2024spl} leverages GPT-4 for Socratic dialogue in a playground-style interface.

On the fine-tuning side, EduChat \citep{dan2023educhat} and SocraticLM \citep{liu2024socraticLM} train models on Socratic-style datasets to internalize heuristic questioning patterns, while \citet{ding2024socraticmath} apply the Socratic method specifically to conversational mathematics teaching. MathDial \citep{macina2023mathdial} contributes a tutoring dialogue dataset grounded in math reasoning with rich pedagogical annotations.

However, as \citet{peng2025kele} observe, these approaches primarily focus on generating heuristic questions at the individual turn level without enforcing coherent multi-stage teaching progression. The absence of structured rules can lead to arbitrary guidance, abrupt topic shifts, and failure to consolidate learning. KELE addresses this through its five-stage SocRule system and dual-agent architecture, making it---to our knowledge---the first framework to systematically implement structured Socratic teaching with LLMs.

\subsection{Multi-Agent Collaborative Systems}

Multi-agent architectures have emerged as an effective paradigm for complex LLM tasks that benefit from separation of concerns \citep{ge2023openagi}. In the educational domain, the key insight of KELE's consultant--teacher design is that teaching \emph{planning} (what stage and strategy to use) and teaching \emph{execution} (generating the actual dialogue) are fundamentally different capabilities that benefit from distinct agents \citep{peng2025kele}. This mirrors real-world educational practice where teaching consultants advise on instructional strategy while classroom teachers handle direct student interaction \citep{nevin2009collaborative}. Single-agent approaches struggle with both responsibilities simultaneously, often producing responses that are individually reasonable but lack coherent pedagogical progression \citep{an2024llmcontext}. Our work investigates whether the consultant role---essentially a state classification task---can be more effectively handled by a specialized classifier rather than a general-purpose LLM.

%% ============================================================
%% SECTION 3 — DATASET & METHODOLOGY
%% ============================================================
\section{Dataset and Methodology}

\subsection{SocratDataset}

SocratDataset \citep{peng2025kele} comprises 6,803 multi-turn Chinese dialogues totalling 42,892 annotated interaction turns, drawn from elementary school science education. Each dialogue simulates a Socratic tutoring session in which a teacher guides a student from an initial question to a correct, self-discovered answer. Every turn is annotated with one of 35 pedagogical states organized across five progressive teaching stages (Table~\ref{tab:socrule}). Dialogues average 6.3 turns (range 5--12). The dataset is split 90/10 into 6,122 training dialogues and 681 test dialogues (seed = 42), following the original paper's partition.

\subsection{SocRule: Five-Stage Teaching Framework}

KELE operationalizes Socratic pedagogy through SocRule, a rule system derived from the teaching science literature \citep{trowbridge2011teaching} that decomposes each dialogue into five strictly ordered stages:

\begin{enumerate}
    \item[\textbf{(a)}] \textbf{Student Questioning} (2 states). The student poses an initial question; the consultant determines whether the question is clear enough to proceed.
    \item[\textbf{(b)}] \textbf{Concept Probing} ($\leq$3 turns, 6 states). The teacher probes the student's prior knowledge and identifies misconceptions through diagnostic questions.
    \item[\textbf{(c)}] \textbf{Inductive Reasoning} ($\leq$5 turns, 22 states). The most complex stage: the teacher guides the student to form hypotheses, test predictions against evidence, and reason inductively toward correct explanations.
    \item[\textbf{(d)}] \textbf{Rule Construction} ($\leq$3 turns, 4 states). The student synthesizes observations into a general rule or definition, with the teacher providing corrective scaffolding as needed.
    \item[\textbf{(e)}] \textbf{Teacher Summary} (1 state). Once the student arrives at the correct answer, the teacher summarizes the key concepts and concludes the dialogue.
\end{enumerate}

Stage progression is strictly monotonic---once the dialogue advances from stage $s_i$ to $s_{i+1}$, it cannot regress. Each state maps to a concrete teaching strategy (e.g., ``provide a counterexample,'' ``ask the student to re-examine prior learning'') that dictates the teacher's pedagogical move. A maximum of 8 teaching rounds per dialogue is enforced; if the limit is reached before stage (e), the system forces a transition to the final rule-construction state (d33) to ensure graceful conclusion.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{@{}llrr@{}}
    \toprule
    \textbf{Stage} & \textbf{Phase} & \textbf{States} & \textbf{Turns} \\
    \midrule
    a & Student Questioning    & 2  & 6{,}803 \\
    b & Concept Probing        & 6  & 7{,}480 \\
    c & Inductive Reasoning    & 22 & 14{,}445 \\
    d & Rule Construction      & 4  & 7{,}361 \\
    e & Teacher Summary        & 1  & 6{,}803 \\
    \midrule
      & \textbf{Total}         & \textbf{35} & \textbf{42{,}892} \\
    \bottomrule
  \end{tabular}
  \caption{SocRule stage overview. \emph{States} = number of distinct teaching states per stage. \emph{Turns} = total annotated turns in SocratDataset. Stage~c dominates with 22 states and 33.7\% of all turns, reflecting the complexity of inductive reasoning in Socratic dialogue.}
  \label{tab:socrule}
\end{table}

\subsection{System Architecture}

KELE employs a consultant--teacher dual-agent architecture in which the two roles operate in sequence on every turn. Given a student utterance $u_t$ and the full dialogue history $H_{<t}$:

\paragraph{Consultant.} The consultant receives $(H_{<t}, u_t)$ along with the complete SocRule specification encoded in its system prompt. It outputs a structured JSON response containing (1)~an \emph{evaluation}: a free-text analysis of the student's current cognitive state, and (2)~a \emph{state code} (e.g., \texttt{c12}) identifying the applicable teaching state. The state code indexes into a deterministic state-to-action mapping that selects the corresponding teaching strategy.

\paragraph{Teacher.} The teacher receives $(H_{<t}, u_t)$ together with the consultant's evaluation and the action string for the selected state. It generates a natural-language response---typically a single Socratic question---that implements the prescribed strategy while maintaining an encouraging, age-appropriate tone.

\paragraph{Enforcement.} Our implementation enforces SocRule constraints programmatically: if the consultant proposes a state from an earlier stage than the current one, the system overrides the regression and retains the current stage. This phase-monotonicity guard ensures strict forward progression regardless of consultant errors.

\subsection{Reproduction Setup}

We reimplement the full KELE pipeline and evaluate on the 681-dialogue test split under the following configuration:

\begin{itemize}
    \item \textbf{Teacher:} SocratTeachLLM (9.4B parameters; GLM4-9B \citep{glm2024chatglm} fine-tuned via LoRA \citep{hu2022lora} on SocratDataset's training split), served locally via vLLM \citep{kwon2023vllm} on an NVIDIA RTX 5090 (32\,GB).
    \item \textbf{Consultant:} GPT-4o-2024-11-20 \citep{openai2024gpt4o} via the OpenAI API, matching the original paper's consultant choice.
    \item \textbf{Inference:} Each test dialogue is run end-to-end through the live system. The consultant produces predictions from scratch on every turn---it does \emph{not} receive ground-truth state labels.
\end{itemize}

\noindent This last point constitutes the most significant methodological difference from the original evaluation. \citet{peng2025kele} feed ground-truth consultant labels directly to the teacher, measuring teacher generation quality in isolation. Our evaluation measures the full pipeline including consultant errors, producing a more realistic assessment of end-to-end system performance. As we show in Table~\ref{tab:reproduction}, this accounts for the majority of the ROUGE gap between our results and the published figures.

\subsection{Evaluation Metrics}

We evaluate along two complementary axes:

\paragraph{Text overlap.} We compute ROUGE-1, ROUGE-2, and ROUGE-L \citep{lin2004rouge} using character-level tokenization (each Chinese character treated as a token), which is the standard convention for Chinese text evaluation. BLEU-4 \citep{papineni2002bleu} is computed with word-level segmentation. These metrics measure how closely the generated teacher responses match the ground-truth responses in the dataset.

\paragraph{State accuracy.} We compare the consultant's predicted state code against the ground-truth annotation for each turn, reporting both overall accuracy and per-stage breakdowns. State accuracy is a critical diagnostic metric: an incorrect state prediction causes the teacher to receive the wrong teaching strategy, which can produce pedagogically incoherent responses even when the teacher model itself is capable. As we report in Section~4, accuracy varies dramatically across stages---from 95\% on the binary stage~(a) to below 6\% on the 22-way classification in stage~(c)---motivating our proposed improvement of replacing the LLM consultant with a supervised classifier.

\subsection{Consultant Sensitivity Analysis}

The \$17.49 cost per evaluation run limits iteration speed. To quantify whether a weaker, locally hosted consultant could substitute, we ran two additional full evaluations: Qwen3.5-9B served via vLLM on WAVE HPC (NVIDIA V100, 32\,GB), and Qwen2.5:7B served via Ollama on an Apple M4 Mac Mini (CPU inference), with teacher inference on an AMD Radeon R9700 (32\,GB, ROCm). Table~\ref{tab:consultant} summarizes all three runs.

\begin{table}[t]
  \centering
  \small
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{@{}lcccccc@{}}
    \toprule
    \textbf{Consultant} & \textbf{R-1} & \textbf{R-2} & \textbf{R-L} & \textbf{B-4} & \textbf{St.\ acc} & \textbf{Cost} \\
    \midrule
    GPT-4o (ours)         & \textbf{44.61} & \textbf{26.04} & \textbf{38.02} & \textbf{19.60} & \textbf{25.94\%} & \$17.49 \\
    Qwen3.5:9B (V100)     & 43.72 & 24.87 & 36.76 & 18.63 & 18.93\% & \$0 \\
    Qwen2.5:7B (M4 CPU)   & 43.57 & 24.90 & 36.91 & 18.56 & 15.16\% & \$0 \\
    \bottomrule
  \end{tabular}%
  }
  \caption{Consultant sensitivity across the 681-dialogue test split. ROUGE/BLEU varies by at most 1.0 R-1 across all runs, while state accuracy drops 11 points as consultant capability decreases from GPT-4o to Qwen2.5:7B. Text-overlap is driven by the fixed SocratTeachLLM teacher; state accuracy is the consultant-sensitive signal.}
  \label{tab:consultant}
\end{table}

The central finding is that ROUGE and BLEU are \emph{insensitive} to consultant quality: a free local 7B model costs only $-$1.0 ROUGE-1 and $-$1.0 BLEU-4 relative to GPT-4o. State accuracy, by contrast, degrades sharply with consultant capability, particularly on stage~c (22 fine-grained inductive-reasoning states), which drops from 4.70\% (GPT-4o) to 0.63\% (Qwen2.5:7B). This confirms that the consultant bottleneck is a structured classification problem poorly suited to zero-shot prompting of general-purpose LLMs.

\subsection{Proposed Improvements}

We target three complementary weaknesses identified in the baseline evaluation.

\paragraph{Hierarchical state classifier.}
We propose replacing the LLM consultant with a supervised two-stage classifier built on a pretrained Chinese BERT encoder \citep{devlin2019bert}. A first classification head predicts the teaching stage (5-way softmax); a second head predicts the within-stage state conditioned on the predicted stage (up to 22-way for stage~c). Both heads are trained jointly on the 42,000 labeled turns of the SocratDataset training split. A 125M-parameter classifier trained directly on in-domain labels should substantially outperform zero-shot LLM prompting on this structured prediction task while eliminating per-run API cost entirely.

\paragraph{Teacher base model upgrade.}
SocratTeachLLM is fine-tuned from GLM4-9B (C-Eval $\approx$81.5), which lags behind more recent Chinese-capable open-weight models. We propose re-running the SocratDataset LoRA fine-tune starting from \texttt{deepseek-ai/DeepSeek-R1-Distill-Qwen-14B} (C-Eval 91.8, MIT license), which inherits chain-of-thought reasoning behavior well-suited to multi-step Socratic dialogue. We will ablate against \texttt{Qwen/Qwen3-14B} (dense architecture, Apache 2.0) and the direct architectural successor of the current base, \texttt{zai-org/GLM-4-9B-0414}, to isolate the contribution of base-model capability from fine-tuning data quality.

\paragraph{Retrieval-augmented teacher.}
To address the surface-form gap that drives the BLEU-4 shortfall, we propose indexing the SocratDataset training split by topic and concept using a Chinese sentence encoder (\texttt{BAAI/bge-large-zh-v1.5}) and prepending the top-3 most similar past dialogues as few-shot exemplars at inference time. Retrieval injects corpus-specific phrasing style without additional fine-tuning and is composable with both the classifier and the model-upgrade improvements.

%% ============================================================
%% SECTION 4 — EVALUATION & RESULTS (placeholder)
%% ============================================================
\section{Evaluation and Results}

% To be completed for the May 5 submission.

%% ============================================================
%% SECTION 5 — CONCLUSION (placeholder)
%% ============================================================
\section{Conclusion}

% To be completed for the May 14 submission.

%% ============================================================
%% LIMITATIONS (placeholder)
%% ============================================================
\section*{Limitations}

% To be completed for the May 14 submission.

%% ============================================================
%% ETHICS (placeholder)
%% ============================================================
\section*{Ethics Statement}

% To be completed for the May 14 submission.

\bibliography{custom}

\end{document}
